---
title: "QUANT_only_summary"
output: github_document
date: "2025-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cowplot)
```

# SIRVs

## Pearson Cor

```{r}

SIRVs_pearson_cor_df = read.csv("SIRVs.pearson_cor.tsv", header=T, sep="\t")  %>% dplyr::select(name, cor_value, data_name)

SIRVs_pearson_cor_df$data_name = str_replace(SIRVs_pearson_cor_df$data_name, "CL_", "")
SIRVs_pearson_cor_df$data_name = str_replace(SIRVs_pearson_cor_df$data_name, "_sirv", "")

SIRVs_pearson_cor_df = SIRVs_pearson_cor_df %>% separate_wider_delim(data_name, "_", names=c("cell_line", "sirv_set"))

SIRVs_pearson_cor_df %>% head()

```


```{r}
# rank according to mean cor val
pearson_cor_df_ranking = SIRVs_pearson_cor_df %>% 
  filter(sirv_set != "E0") %>%
  group_by(name) %>% summarize(mcor = mean(cor_value)) %>% arrange(desc(mcor)) %>% pull(name)


SIRVs_pearson_cor_df$name = factor(SIRVs_pearson_cor_df$name, levels=pearson_cor_df_ranking)

```

```{r}

SIRVs_pearson_cor_df %>%
  filter(sirv_set != "E0") %>%
  ggplot(aes(x=name, y=cor_value)) + geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_grid(vars(cell_line), vars(sirv_set)) +
  ggtitle("Pearson correlation")


```


## Spearman Cor

```{r}

SIRVs_spearman_cor_df = read.csv("SIRVs.spearman_cor.tsv", header=T, sep="\t")  %>% dplyr::select(name, cor_value, data_name)


SIRVs_spearman_cor_df$data_name = str_replace(SIRVs_spearman_cor_df$data_name, "CL_", "")
SIRVs_spearman_cor_df$data_name = str_replace(SIRVs_spearman_cor_df$data_name, "_sirv", "")

SIRVs_spearman_cor_df = SIRVs_spearman_cor_df %>% separate_wider_delim(data_name, "_", names=c("cell_line", "sirv_set"))

SIRVs_spearman_cor_df %>% head()


```


```{r}
# rank according to mean cor val
spearman_cor_df_ranking = SIRVs_spearman_cor_df %>% 
  filter(sirv_set != "E0") %>%
  group_by(name) %>% summarize(mcor = mean(cor_value)) %>% arrange(desc(mcor)) %>% pull(name)


SIRVs_spearman_cor_df$name = factor(SIRVs_spearman_cor_df$name, levels=pearson_cor_df_ranking)

```

```{r}

SIRVs_spearman_cor_df %>% 
     filter(sirv_set != "E0") %>%
  ggplot(aes(x=name, y=cor_value)) + geom_col() +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_grid(vars(cell_line), vars(sirv_set)) +
  ggtitle("SIRVs - Spearman correlation")


```


## Median relative difference

```{r}
SIRVs_median_rel_diff = read.csv("SIRVs.median_rel_diff.tsv", sep="\t", header=T) %>%
  dplyr::select(name, rel_diff, data_name)

SIRVs_median_rel_diff$data_name = str_replace(SIRVs_median_rel_diff$data_name, "CL_", "")
SIRVs_median_rel_diff$data_name = str_replace(SIRVs_median_rel_diff$data_name, "_sirv", "")

SIRVs_median_rel_diff = SIRVs_median_rel_diff %>% separate_wider_delim(data_name, "_", names=c("cell_line", "sirv_set"))

SIRVs_median_rel_diff %>% head()
```


```{r}
# rank according to median relative difference
mrel_median_diff_ranking = SIRVs_median_rel_diff %>% group_by(name) %>% summarize(mrel_diff = mean(rel_diff)) %>% arrange(mrel_diff) %>% pull(name)


SIRVs_median_rel_diff$name = factor(SIRVs_median_rel_diff$name, levels=mrel_median_diff_ranking)

```


```{r}

SIRVs_median_rel_diff %>% ggplot(aes(x=name, y=rel_diff)) + geom_col() +
    theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   facet_grid(vars(cell_line), vars(sirv_set)) +
   ggtitle("SIRVs - median relative difference")

```


```{r}

SIRVs_median_rel_diff %>% ggplot(aes(x=name, y=rel_diff, fill=cell_line)) + 
   geom_bar(stat='identity', position='dodge') +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   ggtitle("SIRVs - median relative difference") +
  facet_wrap(~sirv_set, nrow=1)


```



```{r}

SIRVs_median_rel_diff %>% group_by(name, sirv_set) %>%
  summarise(mean_median_rel_diff = mean(rel_diff),
            stderr = sd(rel_diff)/sqrt(n())) %>% 
  ungroup()  %>%
  ggplot() + 
  geom_col(aes(x=name, y=mean_median_rel_diff)) +
  geom_errorbar(aes(x = name, ymin = mean_median_rel_diff - stderr, ymax = mean_median_rel_diff + stderr)) +
  facet_wrap(~sirv_set) +
  theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  ggtitle("SIRVs - mean medial rel diff")
  

```



## Mean relative difference


```{r}
SIRVs_mean_rel_diff = read.csv("SIRVs.mean_rel_diff.tsv", sep="\t", header=T) %>%
  dplyr::select(name, rel_diff, data_name)

SIRVs_mean_rel_diff$data_name = str_replace(SIRVs_mean_rel_diff$data_name, "CL_", "")
SIRVs_mean_rel_diff$data_name = str_replace(SIRVs_mean_rel_diff$data_name, "_sirv", "")

SIRVs_mean_rel_diff = SIRVs_mean_rel_diff %>% separate_wider_delim(data_name, "_", names=c("cell_line", "sirv_set"))

SIRVs_mean_rel_diff %>% head()
```


```{r}
# rank according to median relative difference
mrel_mean_diff_ranking = SIRVs_mean_rel_diff %>% group_by(name) %>% summarize(mrel_diff = mean(rel_diff)) %>% arrange(mrel_diff) %>% pull(name)


SIRVs_mean_rel_diff$name = factor(SIRVs_mean_rel_diff$name, levels=mrel_mean_diff_ranking)

```




```{r}


SIRVs_mean_rel_diff %>% ggplot(aes(x=name, y=rel_diff)) + geom_col() +
  theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   facet_grid(vars(cell_line), vars(sirv_set)) +
   ggtitle("SIRVs - mean relative difference")




```




```{r}

SIRVs_mean_rel_diff %>% ggplot(aes(x=name, y=rel_diff, fill=cell_line)) +
  theme_bw() +
   geom_bar(stat='identity', position='dodge') +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   ggtitle("SIRVs - mean relative difference")


```

```{r}

get_latest_programs = function(data_df) {

  program_versions = data_df %>%
    distinct(name) %>%
    mutate(name = as.character(name),
           Tool = str_remove(name, "-v[0-9].*$"),
           Version = str_match(name, "-v([0-9][0-9A-Za-z.\\-]*)$")[, 2])

  latest_versioned_programs = program_versions %>%
    filter(! is.na(Version)) %>%
    group_by(Tool) %>%
    arrange(desc(numeric_version(Version)), .by_group = TRUE) %>%
    slice(1) %>%
    ungroup()

  unversioned_programs = program_versions %>%
    filter(is.na(Version))

  bind_rows(latest_versioned_programs, unversioned_programs) %>%
    pull(name)
}

latest_programs = get_latest_programs(SIRVs_mean_rel_diff)

latest_programs

```


```{r}

SIRVs_mean_rel_diff %>% group_by(name, sirv_set) %>%
  summarise(mean_mean_rel_diff = mean(rel_diff),
            stderr = sd(rel_diff)/sqrt(n())) %>% 
  ungroup()  %>%
  ggplot() + 
  geom_col(aes(x=name, y=mean_mean_rel_diff)) +
  geom_errorbar(aes(x = name, ymin = mean_mean_rel_diff - stderr, ymax = mean_mean_rel_diff + stderr)) +
  facet_wrap(~sirv_set) +
  theme_bw() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  ggtitle("SIRVs - mean mean rel diffs")
  

```

# heatmap style representation

```{r}

SIRVs_mean_mean_rel_diff  = SIRVs_mean_rel_diff %>% group_by(name, sirv_set) %>%
  summarise(mean_mean_rel_diff = mean(rel_diff))

SIRVs_mean_mean_rel_diff %>% head()
```



```{r}

get_ranked_df = function(sirv_set_val) {
  
  ranked_df =   SIRVs_mean_mean_rel_diff %>% filter(sirv_set == sirv_set_val)
  
  ranked_progs = ranked_df %>% arrange(mean_mean_rel_diff) %>% pull(name)
  
  ranked_df$name = factor(ranked_df$name, levels=ranked_progs)
  
  return(ranked_df)
 
}

```



```{r}

color_scale_limit_range = c(0,1.05)

```


```{r}

E0_mean_expr_df = get_ranked_df("E0")

E0_mean_expr_plot = E0_mean_expr_df %>% ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) + 
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E0 mean rel diff") +
  ylab("Mean Relative Difference")

E0_mean_expr_plot

```


```{r}

E1_mean_expr_df = get_ranked_df("E1")

E1_mean_expr_plot = E1_mean_expr_df %>% 
  ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) + 
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E1 mean rel diff") +
  ylab("Mean Relative Difference")

E1_mean_expr_plot

```

```{r}

E2_mean_expr_df = get_ranked_df("E2")

E2_mean_expr_plot = E2_mean_expr_df %>% 
  ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) + 
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E2 mean rel diff") +
  ylab("Mean Relative Difference")

E2_mean_expr_plot

```


```{r}

library(cowplot)

summary_plot = plot_grid(E0_mean_expr_plot,
                              E1_mean_expr_plot,
                              E2_mean_expr_plot,
                              ncol=1) 

summary_plot

```





```{r}

ggsave(summary_plot, file="SIRVs_sim_quant_summary.pdf", width=8, height=10.5)

```

```{r}

SIRVs_latest_mean_mean_rel_diff = SIRVs_mean_rel_diff %>%
  filter(as.character(name) %in% latest_programs) %>%
  group_by(name, sirv_set) %>%
  summarise(mean_mean_rel_diff = mean(rel_diff))

get_ranked_latest_df = function(sirv_set_val) {

  ranked_df = SIRVs_latest_mean_mean_rel_diff %>%
    filter(sirv_set == sirv_set_val)

  ranked_progs = ranked_df %>%
    arrange(mean_mean_rel_diff) %>%
    pull(name)

  ranked_df$name = factor(ranked_df$name, levels=ranked_progs)

  return(ranked_df)
}

E0_latest_mean_expr_df = get_ranked_latest_df("E0")

E0_latest_mean_expr_plot = E0_latest_mean_expr_df %>%
  ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) +
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E0 mean rel diff") +
  ylab("Mean Relative Difference")

E1_latest_mean_expr_df = get_ranked_latest_df("E1")

E1_latest_mean_expr_plot = E1_latest_mean_expr_df %>%
  ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) +
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E1 mean rel diff") +
  ylab("Mean Relative Difference")

E2_latest_mean_expr_df = get_ranked_latest_df("E2")

E2_latest_mean_expr_plot = E2_latest_mean_expr_df %>%
  ggplot(aes(x=name, y=mean_mean_rel_diff, fill=mean_mean_rel_diff)) +
  geom_col() +
  geom_text(aes(label = round(mean_mean_rel_diff, 2)), vjust=-0.5, size = 3, color = "black") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen", limits=color_scale_limit_range) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("E2 mean rel diff") +
  ylab("Mean Relative Difference")

latest_summary_plot = plot_grid(E0_latest_mean_expr_plot,
                                E1_latest_mean_expr_plot,
                                E2_latest_mean_expr_plot,
                                ncol=1)

latest_summary_plot

```

```{r}

ggsave(latest_summary_plot,
       file="SIRVs_sim_quant_summary.latest_tools.pdf",
       width=8,
       height=10.5)

```

## Oarfish-style quantification metrics

```{r oarfish_read_metrics}

SIRVs_oarfish_metrics_file = "SIRVs.oarfish_style_metrics.tsv"

if (file.exists(SIRVs_oarfish_metrics_file)) {
  SIRVs_oarfish_metrics_df = read.csv(SIRVs_oarfish_metrics_file, sep="\t", header=T)
  SIRVs_oarfish_metrics_df$data_name = str_replace(SIRVs_oarfish_metrics_df$data_name, "CL_", "")
  SIRVs_oarfish_metrics_df$data_name = str_replace(SIRVs_oarfish_metrics_df$data_name, "_sirv", "")
  SIRVs_oarfish_metrics_df = SIRVs_oarfish_metrics_df %>%
    separate_wider_delim(data_name, "_", names=c("cell_line", "sirv_set"))
} else {
  message(SIRVs_oarfish_metrics_file, " not found; run make summarize after regenerating per-sample oarfish_style_metrics.tsv files.")
  SIRVs_oarfish_metrics_df = tibble()
}

SIRVs_oarfish_metrics_df %>% head()

```

```{r oarfish_prep_metrics}

higher_is_better_metrics = c(
  "spearman",
  "pearson_log1p",
  "ccc_log1p",
  "kendall",
  "pr_auc",
  "average_precision"
)

lower_is_better_metrics = c(
  "rmse",
  "nrmse_mean_truth",
  "mard"
)

oarfish_metric_labels = tribble(
  ~metric, ~metric_label,
  "spearman", "Spearman",
  "pearson_log1p", "Pearson log1p",
  "ccc_log1p", "CCC log1p",
  "kendall", "Kendall",
  "rmse", "RMSE",
  "nrmse_mean_truth", "NRMSE",
  "mard", "MARD",
  "pr_auc", "PR-AUC",
  "average_precision", "Average precision"
)

if (nrow(SIRVs_oarfish_metrics_df) > 0) {
  SIRVs_oarfish_metrics_long = SIRVs_oarfish_metrics_df %>%
    dplyr::select(name, cell_line, sirv_set, all_of(c(higher_is_better_metrics, lower_is_better_metrics))) %>%
    pivot_longer(
      cols = all_of(c(higher_is_better_metrics, lower_is_better_metrics)),
      names_to = "metric",
      values_to = "metric_value"
    ) %>%
    left_join(oarfish_metric_labels, by="metric") %>%
    mutate(
      data_name = paste(cell_line, sirv_set, sep="_"),
      direction = if_else(metric %in% higher_is_better_metrics, "higher is better", "lower is better"),
      metric_label = factor(metric_label, levels=oarfish_metric_labels$metric_label)
    )
} else {
  SIRVs_oarfish_metrics_long = tibble()
}

```

```{r oarfish_rank_metrics}

if (nrow(SIRVs_oarfish_metrics_long) > 0) {
  SIRVs_oarfish_metric_rankings = SIRVs_oarfish_metrics_long %>%
    group_by(data_name, metric, metric_label, direction) %>%
    mutate(
      metric_rank = if_else(
        direction == "higher is better",
        min_rank(desc(metric_value)),
        min_rank(metric_value)
      )
    ) %>%
    ungroup()

  SIRVs_oarfish_mean_rankings = SIRVs_oarfish_metric_rankings %>%
    group_by(name, metric, metric_label, direction) %>%
    summarize(
      mean_metric_value = mean(metric_value, na.rm=TRUE),
      mean_rank = mean(metric_rank, na.rm=TRUE),
      .groups = "drop"
    )
} else {
  SIRVs_oarfish_metric_rankings = tibble()
  SIRVs_oarfish_mean_rankings = tibble()
}

```

```{r oarfish_reorder_helpers}

reorder_within = function(x, by, within, fun=mean, sep="___", ...) {
  new_x = paste(x, within, sep=sep)
  stats::reorder(new_x, by, FUN=fun)
}

scale_y_reordered = function(..., sep="___") {
  scale_y_discrete(labels=function(x) gsub(paste0(sep, ".+$"), "", x), ...)
}

```

Note: PR-AUC and average precision require both expressed and unexpressed truth
isoforms in the scored reference set. In the current SIRV truth sets, all scored
multi-exon SIRV isoforms are truth-expressed (`n_truth_unexpressed = 0`), so
those two metrics are reported as `NA` and their plot facets are not
interpretable for SIRVs.

```{r oarfish_mean_rank_plot}

if (nrow(SIRVs_oarfish_mean_rankings) > 0) {
  oarfish_metric_mean_rank_plot = SIRVs_oarfish_mean_rankings %>%
    ggplot(aes(x=mean_rank, y=reorder_within(name, -mean_rank, metric_label), fill=mean_rank)) +
    geom_col() +
    geom_text(aes(label=round(mean_rank, 1)), hjust=-0.1, size=2.5) +
    facet_wrap(~metric_label, scales="free_y") +
    scale_y_reordered() +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    scale_fill_gradient(low="darkgreen", high="lightgreen") +
    theme_bw() +
    theme(axis.text.y = element_text(size=6),
          legend.position="none") +
    xlab("Mean rank across SIRV samples") +
    ylab("") +
    ggtitle("SIRVs - Oarfish-style metric rankings")

  oarfish_metric_mean_rank_plot
}

```

```{r oarfish_mean_value_plot}

if (nrow(SIRVs_oarfish_mean_rankings) > 0) {
  oarfish_metric_value_plot = SIRVs_oarfish_mean_rankings %>%
    ggplot(aes(x=mean_metric_value, y=reorder_within(name, mean_metric_value, metric_label), fill=mean_metric_value)) +
    geom_col() +
    facet_wrap(~metric_label, scales="free") +
    scale_y_reordered() +
    theme_bw() +
    theme(axis.text.y = element_text(size=6),
          legend.position="none") +
    xlab("Mean metric value across SIRV samples") +
    ylab("") +
    ggtitle("SIRVs - Oarfish-style mean metric values")

  oarfish_metric_value_plot
}

```

```{r oarfish_save_plots}

if (exists("oarfish_metric_mean_rank_plot")) {
  ggsave(oarfish_metric_mean_rank_plot,
         file="SIRVs_oarfish_style_metric_rankings.pdf",
         width=11,
         height=8.5)
}

if (exists("oarfish_metric_value_plot")) {
  ggsave(oarfish_metric_value_plot,
         file="SIRVs_oarfish_style_metric_values.pdf",
         width=11,
         height=8.5)
}

```

```{r oarfish_latest_rank_plot}

if (nrow(SIRVs_oarfish_mean_rankings) > 0) {
  SIRVs_oarfish_latest_mean_rankings = SIRVs_oarfish_mean_rankings %>%
    filter(as.character(name) %in% latest_programs)

  oarfish_latest_metric_mean_rank_plot = SIRVs_oarfish_latest_mean_rankings %>%
    ggplot(aes(x=mean_rank, y=reorder_within(name, -mean_rank, metric_label), fill=mean_rank)) +
    geom_col() +
    geom_text(aes(label=round(mean_rank, 1)), hjust=-0.1, size=2.5) +
    facet_wrap(~metric_label, scales="free_y") +
    scale_y_reordered() +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    scale_fill_gradient(low="darkgreen", high="lightgreen") +
    theme_bw() +
    theme(axis.text.y = element_text(size=6),
          legend.position="none") +
    xlab("Mean rank across SIRV samples") +
    ylab("") +
    ggtitle("SIRVs - Oarfish-style metric rankings, latest tools")

  oarfish_latest_metric_mean_rank_plot
}

```

```{r oarfish_save_latest_plot}

if (exists("oarfish_latest_metric_mean_rank_plot")) {
  ggsave(oarfish_latest_metric_mean_rank_plot,
         file="SIRVs_oarfish_style_metric_rankings.latest_tools.pdf",
         width=11,
         height=8.5)
}

```
